Animacy Detection in Stories

### 2015-01-20 paper/paper.tex
updates

\usepackage{gb4e}
\bibliographystyle{plain}

@@

\title{Animacy Detection in Stories}
\author[1]{Folgert Karsdorp}
\author[2]{Marten van der Meulen}
\author[3]{Theo Meder}
\author[4]{Antal van den Bosch}
\affil[1]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{folgert.karsdorp@meertens.knaw.nl}}
\affil[2]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{marten.van.der.meulen@meertens.knaw.nl}}
\affil[3]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{theo.meder@meertens.knaw.nl}}
\affil[4]{Radboud University\\
  Nijmegen, The Netherlands\\
  \texttt{a.vandenbosch@let.ru.nl}}
%\authorrunning{F. Karsdorp and M. van der Meulen and A. van den Bosch}
%\Copyright{Folgert Karsdorp and Marten van der Meulen and Antal van
%  den Bosch}
\authorrunning{F. Karsdorp and M. van der Meulen and T. Meder and A. van den Bosch}
\Copyright{Folgert Karsdorp and Marten van der Meulen and Theo Meder
  and Antal van den Bosch}
\subjclass{I.2.7 Natural Language Processing}
\keywords{animacy detection, deep learning, folklore}

@@

\begin{abstract}
  Existing animacy detection systems, while scoring well, seem to be
  out of sync with current (psycho)linguistic research, which takes a
  more dynamic approach to animacy. This paper presents a
  linguistically uninformed computational model for animacy
  classification. The model makes use of word $n$-grams in
  combinations with lower dimensional word embedding representations
  that were learned from a web-scale corpus. We compare the model to a
  number of linguistically informed models that use features such as
  dependency tags and show competitive results.
\end{abstract}

\section{Introduction}

@@

Traditionally, even though animacy may be a scalar rather than a
strictly categorical distinction (see e.g. the animacy hierarchy in
\cite{comrie:89} and research such as \cite{rosenbach:08}), the
animate/inanimate distinction is taken as fixed with regard to lexical
items. This standpoint has been challenged, however, by researchers
from different fields. Firstly, it has long been established in
linguistic typology that not all languages award animacy to the same
entities in different grammatical categories: as \cite{comrie:89} for
example notes, English, and many other languages, distinguishes
between human and not-human in the choice of pronouns, whereas Russian
distinguishes between animate, entailing humans and animals, versus
non-animate, in their interrogative pronouns. This indicates different
subdivisions of animacy in the respective languages. Secondly,
philosophers such as Daniel Dennett support the view that animacy and
aliveness are to be treated as epistemological stances rather than
fixed states in the world: not ineffable qualia but behavioral
capacity defines our stance towards objects \cite{dennett:96}. In
other words, depending on whether people \textit{think} that an object
is animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Finally, evidence from
psycholinguistic research has accumulated to support the view of
animacy as a cognitive viewpoint rather than an extra-perceptive
absolute. \cite{nieuwland:05} for example show that test subjects
readily accept animate behavior from inanimate objects within the
proper contexts, and \cite{vogels:13} moreover emphasize the relation
between animacy and motion, showing that factors such as
self-propelment play a crucial role in recognizing and/or awarding
animacy to certain objects. This is exemplified in the opening of this
well-known story:

@@

In this paper we present a new linguistically uninformed model to
automatically label texts for animacy. We show that we can (mostly) do
away with features that require syntactic parsing or semantic lexicons
while still yielding competitive performance. We focus on labeling
animacy in stories because stories pose some interesting problems to
automatic systems of animacy recognition. As the example of the
fleeing pancake already made clear, quite some entities in stories,
though inanimate in the `real' world, exhibit all behavioral
capacities to consider them animate and rational agents. Another
example is the \textit{Sorcerer's Apprentice} sequence in Walt
Disney's famous \textit{Fantasia}, in which brooms display the ability
to collect buckets of water. Such examples, in which otherwise as
inanimatedly labeled entities such as pancakes and brooms act, make a
clear case for developing dynamic, data driven systems that do not
rely too much on static and fixed world knowledge, but rather on
immediate context.

The rest of this paper is structured as follows. We will start with a
short overview of existing techniques for automatically labeling
animacy in texts, including the definitions of animacy used in these
papers (\S\ref{sec:previous-work}). After a description of the corpus
used in our study and how the annotations of the corpus have been
established (\S\ref{sec:data}), we will give an account of our
computational models in Section \ref{sec:models}. We report on the
empirical results in Section \ref{sec:results}. The last section
offers our conclusions and possible directions for future research.

@@

To develop this dynamic, data driven system we used a corpus of Dutch
folktales. One of the reasons to use folk tales was that, as
\cite{vogels:13} note, `In cartoons or fairy tales [\ldots] inanimate
entities or animals are often anthropomorphized', which means that the
material could yield interesting cases of unexpected animacy, as is
the case with the pancake in the abovementioned story. Our corpus
consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen}, compiled by
\cite{sinninghe:78}. The collection is made up of Dutch retellings of
popular and widespread stories, including such tales as \textit{The
  Bremen Town Musicians} (ATU 130) and \textit{The Table, the Ass, and
  the Stick } (ATU 563), as well as lesser-known stories such as
\textit{The Singing Bone} (ATU 780) and \textit{Cock, Hen, Duck, Pin,
  and Needle on a Journey} (ATU 210). This last story is again a clear
example where otherwise inanimate objects are animated in a folk
story, as it concerns the adventures of several household items, such
as a \textit{pin}, a \textit{hackle}, an \textit{egg}, and a
\textit{whetstone}. A digital version of the collection is available
in the Dutch Folktale Database from the Meertens Institute (corpus
SINVSUNV.20E).\footnote{See \url{http://www.verhalenbank.nl}} Using a
single collection for our corpus presents us with a helpful
homogeneity with regard to the editor, length of the stories (which is
between X and Y words) and language use. On the other hand, the
collection displays diversity as well in the choice of the stories,
which entail fairy tales, legends, and sagas.

@@

This annotation resulted in 11542 animate tokens of 743 types, while
implicitly yielding 62926 inanimate tokens from 5011 unique inanimate
words. Because of our context-dependent approach, some words, such as
\textit{pancake} and \textit{egg}, occurred in both animate types as
inanimate types, because they were labeled as both animate and
inanimate in some cases in our corpus. It is telling that of the
animate tokens 4627 (40,1\%) were nouns, while only 6878 of the
inanimate tokens (11 \%) are nouns. This shows that being a noun is
already somewhat of an indication for animacy.

After tokenization with the tokenization module of the Python software
package Pattern \cite{smedt:12} we fed all stories to the
state of the art syntactic parser for Dutch, Alpino \cite{bouma:01}.
%dit kan iets uitgebreider eventueel%

@@

\subsection{Task description}
%misschien deze subsection invoegen in bovenstaande inleiding. FK, nee
%ik vind dit altijd een erg prettige plaats (eventueel een
%soortgelijke zin in de inleiding, maar hier mag-ie niet weg :-))
We formulate the problem of animacy detection as a classification
problem where the goal is to assign a label at word level, rather than
at lemma level. This label indicates whether the word is classified as
animate or inanimate.

@@

Animate words are far outnumbered by inanimate words in our collection
(see \S\ref{sec:data}). Reporting accuracy scores would therefore
provide skewed results, strongly favoring the majority category. The
relative rarity of animate words makes evaluation measures such as the
well-known $F1$-score more appropriate. For all experiments we report
on the precision, recall and $F1$-score~\cite{rijsbergen:79}. As a
second measure of evaluation, we compute the Area Under the Precision
Recall Curve (AUC_{PR}) which also takes into account possible class
imbalance. Also, while in most of the literature on animacy detection
results are only presented for the classification of nouns or noun
phrases, we will present these while furthermore reporting on the
results for all words in a text.
%bovenstaande zin loopt nog niet helemaal lekker%

@@

\subsubsection{Semantic Features} The most innovative feature we have
included in our model is concerned with semantic similarity. In his
\textit{Philophische Untersuchungen} Wittgenstein already suggests
that ``Die Bedeutung eines Wortes ist sein Gebrauch in der Sprache''
(PI 43). The well-known insight in computational linguistics that the
meaning of words can be approximated by comparing the linguistic
contexts in which words appear, is an implementation of this very
idea. The idea is very simple, yet powerful: words that often
co-appear with the same set of words, will have a more similar
meaning. Recently, there has been a lot of interest in procedures that
can automatically induce so-called `word embeddings' from large,
unannotated collections of texts
(e.g.~\cite{mikolov:13,pennington:14}). These models typically attempt
to learn a lower dimensional vector for each word in the vocabulary
which captures the typical co-occurrence patterns of the word in the
corpus. The similarity between words can then be approximated by
applying similarity metrics, such as the cosine metric, to these
vectors of word embeddings.

We have trained word embeddings with 300 dimensions using the popular
skip-gram architecture \cite{mikolov:13} on the Dutch corpus of COW
(COrpora from the Web).\footnote{The word embedding vectors have been
  published online at XXX.} COW is a collection of linguistically
processed web corpora for English, Dutch, Spanish, French, Swedish and
German \cite{schaefer:12}. The 2014 Dutch corpus contains 6.8 billion
word tokens. The idea behind using the word embeddings is that
similarities between animate words can be estimated by inspecting the
context in which they occur. From this follows for example that the word
embeddings of an animate word are more similar to those of other
animate words, as opposed to the embeddings of inanimate words. To give an
example, the 5 nearest neighbors of \textit{Assepoester}
(`Cinderella') are: \textit{Sneeuwwitje} (`Snow White') (sim = 0.807),
\textit{Roodkapje} (`Little Red Riding Hood') (sim = 0.747),
\textit{Doornroosje} (`Sleeping Beauty') (sim = 0.715),
\textit{Repelsteeltje} (`Rumpelstiltskin') (sim = 0.669) and
\textit{Alladin} (0.645). In contrast, the 5 most similar words to
\textit{boek} (`book') are: \textit{boekje} (`book-\textsc{dim}')
(sim=0.8), \textit{kinderboek} (`children's book') (0.757),
\textit{stripboek} (`comic') (0.713), \textit{kortverhaal} (`short
story') (0.709), \textit{jeugdboek} (`youth book') (0.706).

\subsection{Models}
We employ a Maximum Entropy classifier with L2 regularization as
implemented in \cite{sklearn}. In all experiments, we set the
regularization strength parameter $C$ to 1.

@@

Although our background corpus is sufficiently large to cover most
words in an unseen text, there will always be rare words for which we
do not have learned word embeddings. Therefore, in order to
effectively make use of the word embedding vectors, we need a way to
deal with out-of-vocabulary items. We adopt a simple strategy were we
make use of a primary classifier and a back-off classifier. For models
6 to 9, we augment each word with its corresponding 300 dimension word
embeddings vector. In the case of out-of-vocabulary words, we resort
to a back-off model that contains all features except the word
embeddings. For example, a model that makes use of words and word
embeddings, will make a prediction on the basis of the word features
alone. In case of the model that solely uses the embeddings (model 5),
the back-off classifier is represented by a majority vote classifier,
which classifies unseen words as inanimate.

@@

In Table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall, $F1$-score and Area Under the Precision-Recall
Curve (AUC_{PR}) for the animate words and the
inanimate words.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class & Precision &  Recall & $F1$ &  AUC_{PR} \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.93 &    0.89 &    0.91 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.96 &    0.99 &    0.98 & \multirow{2}{*}{0.93} \\
                                                       & animate     &       0.94 &    0.78 &    0.85 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.98 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.90 &    0.91 &  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.95} \\
                                                       & animate     &       0.94 &    0.86 &    0.89 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.98 &    0.99 &    0.99 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings}   & animate     &       0.94 &    0.91 &    0.93 & \multirow{-2}{*}{0.97} \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate    &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.91 &    0.93 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.97 &    0.99 &    0.98 & \multirow{2}{*}{0.96} \\
                                                       & animate     &       0.94 &    0.86 &    0.90 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.98 &    0.99 &    0.99 & \multirow{2}{*}{0.97} \\
                                                       & animate     &       0.94 &    0.92 &    0.93 &  \\
\bottomrule
\end{tabular}
\caption{Precision, Recall, $F1$-score and Area Under the Precision Recall Curve (AUC_{PR}) scores for animate and inanimate classes per feature setting for all words.}
\label{tab:results-all}
\end{table}

@@

The model that only uses word embedding features is one of the best
performing models. This is a context insensitive model that operates
on the level of the vocabulary, which means that it will predict the
same outcome for each token of a particular word type. The high
precision and high recall show us that this model has acquired
knowledge about which words \textit{typically} group with animate
words and which with inanimate words. However, the models that combines
the word embeddings with the context sensitive features, such as word
$n$-grams or Part-of-Speech tags, outperform the context insensitive
model.

The best performance is achieved by the model that combines the word
features, part-of-speech tags and the word embeddings, which has an
$F1$-score of 0.93 on animate words and 0.99 on inanimate
words. Adding more features does not result in any more performance
gain.
%misschien hier dan nog een voorbeeld noemen, of verwijzen naar relevant cijfer in de tabel%
Table \ref{tab:results-noun} presents the results for all nouns and
names in the data set. The best performance is again achieved by the
model that combines the word features with the part-of-speech tags and
word embeddings, resulting in an $F1$-score of 0.92 for animate
instances and 0.95 for inanimate instances.

\begin{table}
\centering
\begin{tabular}{llrrrr}
\toprule
features & class &  Precision &  Recall & $F1$ &  AUC_{PR} \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.90 &    0.96 &    0.92 & \multirow{2}{*}{0.91} \\
                                                       & animate       &       0.93 &    0.85 &    0.89 &  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.78 &    0.98 &    0.87 & \multirow{2}{*}{0.91} \\
                                                       & animate       &       0.96 &    0.60 &    0.74 &  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.90 &    0.97 &    0.93 & \multirow{2}{*}{0.96} \\
                                                       & animate       &       0.95 &    0.85 &    0.90 &  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.86 &    0.96 &    0.90 & \multirow{2}{*}{0.94} \\
                                                       & animate       &       0.93 &    0.78 &    0.84 &  \\
\rowcolor{Gray}                                        & inanimate     &       0.93 &    0.96 &    0.95 &  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & animate       &       0.95 &    0.90 &    0.92 &\multirow{-2}{*}{0.97}  \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.94 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.93 &    0.96 &    0.94 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.95 &    0.89 &    0.92 &  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.87 &    0.96 &    0.91 & \multirow{2}{*}{0.95} \\
                                                       & animate       &       0.93 &    0.80 &    0.86 &  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.93 &    0.96 &    0.95 & \multirow{2}{*}{0.97} \\
                                                       & animate       &       0.95 &    0.90 &    0.92 &  \\
\bottomrule
\end{tabular}
\caption{Precision, Recall, $F1$ score and Area Under the Precision Recall Curve (AUC_{PR}) scores for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}

\section{Animacy Detection in the Wild}

In a classical evaluation setup -- as with our approach -- general
practice is to divide an annotated data set into a set for training a
computational system of which the performance is evaluated on a
held-out test set. Even if we include a separate development set, this
setup always runs the risk over overfitting the data. Our annotated
corpus contains a reasonably diverse set of stories, yet it is fairly
small and stylistically homogeneous. Even though we performed a
cross-validation experiment, the chances of overfitting are
considerably high. In this section we investigate how well our animacy
detection model performs on a collection of unannotated texts.

>>>>

\usepackage{gb4e, rotating}
\bibliographystyle{plain}

@@

\title{Animacy Detection in Stories}
\author[1]{Folgert Karsdorp}
\author[2]{Marten van der Meulen}
\author[3]{Theo Meder}
\author[4]{Antal van den Bosch}
\affil[1]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{fbkarsdorp@fastmail.nl}}
\affil[2]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{marten.van.der.meulen@meertens.knaw.nl}}
\affil[3]{Meertens Institute\\
  Amsterdam, The Netherlands\\
  \texttt{theo.meder@meertens.knaw.nl}}
\affil[4]{Radboud University\\
  Nijmegen, The Netherlands\\
  \texttt{a.vandenbosch@let.ru.nl}}
%\authorrunning{F. Karsdorp and M. van der Meulen and A. van den Bosch}
%\Copyright{Folgert Karsdorp and Marten van der Meulen and Antal van
%  den Bosch}
\authorrunning{F. Karsdorp and M. van der Meulen and T. Meder and A. van den Bosch}
\Copyright{Folgert Karsdorp and Marten van der Meulen and Theo Meder
  and Antal van den Bosch}
\subjclass{I.2.7 Natural Language Processing}
\keywords{animacy detection, deep learning, folklore}

@@

\begin{abstract}
  Existing animacy detection systems, while scoring well, seem to be
  out of sync with current (psycho)linguistic research, which takes a
  more dynamic approach to animacy. This paper presents a
  linguistically uninformed computational model for animacy
  classification. This model makes use of word $n$-grams in
  combinations with lower dimensional word embedding representations
  that were learned from a web-scale corpus. We compare the model to a
  number of linguistically informed models that use features such as
  dependency tags and show competitive results.
\end{abstract}
%aanpassen op basis van embedding verhalenbank-evaluatie%
\section{Introduction}

@@

Although animacy may be a scalar rather than a
strictly categorical distinction (see e.g. the animacy hierarchy in
\cite{comrie:89} and research such as \cite{rosenbach:08}), the
animate/inanimate distinction is traditonally taken as fixed with regard to lexical
items: something is either animate (e.g. a human) or not (e.g. a chair). This standpoint has been challenged, however, by researchers
from different fields. Firstly, it has long been established in
linguistic typology that not all languages award animacy to the same
entities in different grammatical categories: as \cite{comrie:89} notes, for
example, English, and many other languages, distinguishes
between human and not-human in the choice of pronouns, whereas Russian
distinguishes between animate (entailing humans and animals) versus
non-animate (entailing everything else) in its interrogative pronouns. This indicates different
subdivisions of animacy in the respective languages. Secondly,
philosophers such as Daniel Dennett support the view that animacy and
aliveness are to be treated as epistemological stances rather than
fixed states in the world: not ineffable qualia but behavioral
capacity defines our stance towards objects \cite{dennett:96}. In
other words, depending on whether people \textit{think} that an object
is animate, they utilize different cognitive strategies to explain and
predict the actions of those objects. Finally, evidence from
psycholinguistic research has accumulated to support this view of
animacy as a cognitive viewpoint rather than an extra-perceptive
absolute. \cite{nieuwland:05} for example show that test subjects
readily accept animate behavior from inanimate objects within the
proper contexts, and \cite{vogels:13} moreover emphasize the relation
between animacy and motion, showing that factors such as
self-propelment play a crucial role in recognizing and/or awarding
animacy to certain objects. This is exemplified in the opening of this
well-known storyhttp://www.verhalenbank.nl/items/show/9636:

@@

In this paper we present a new linguistically uninformed model to
automatically label texts for animacy. We show that we can (mostly) do
away with features that require syntactic parsing or semantic lexicons
while still yielding competitive performance. We focus on labeling
animacy in stories because stories pose some interesting problems to
automatic systems of animacy recognition. As the example of the
fleeing pancake already made clear, in stories quite some entities can exhibit all behavioral
capacities to consider them animate and rational agents, even though they are inanimate in the `real' world. Another
example is the \textit{Sorcerer's Apprentice} sequence in Walt
Disney's famous \textit{Fantasia}, in which brooms display the ability
to collect buckets of water. Such examples, in which otherwise as
inanimatedly labeled entities such as pancakes and brooms act as animate beings, make a
clear case for developing dynamic, data driven systems that do not
rely too much on static and fixed world knowledge, but rather on
immediate context.

The rest of this paper is structured as follows. We will start with a
short overview of existing techniques for automatically labeling
animacy in texts, including the definitions of animacy used in these
papers (\S\ref{sec:previous-work}). After a description of the corpus
used in our study and how the annotations of the corpus have been
established (\S\ref{sec:data}), we will give an account of our
computational models in Section \ref{sec:models}. We report on the
empirical results in Section \ref{sec:results}. Next, we provide an evaluation on a larger dataset, while also showing a real world application of our animacy detection system \ref{sec:listing}.
The last section offers our conclusions and possible directions for future research.

@@

To develop this dynamic, data driven system we use a corpus of Dutch
folktales. One of the reasons to use folk tales is that, as
\cite{vogels:13} note, `In cartoons or fairy tales [\ldots] inanimate
entities or animals are often anthropomorphized', which means that the
material could yield interesting cases of unexpected animacy, as is
the case with the pancake in the abovementioned story.

Our initial corpus consists of 74 Dutch stories from the collection
\textit{Volkssprookjes uit Nederland en Vlaanderen}, compiled by
\cite{sinninghe:78}. The collection is made up of Dutch retellings of
popular and widespread stories, including such tales as \textit{The
  Bremen Town Musicians} (ATU 130) and \textit{The Table, the Ass, and
  the Stick } (ATU 563), as well as lesser-known stories such as
\textit{The Singing Bone} (ATU 780) and \textit{Cock, Hen, Duck, Pin,
  and Needle on a Journey} (ATU 210). This last story is again a clear
example where otherwise inanimate objects are animated in a folk
story, as it concerns the adventures of several household items, such
as a \textit{pin}, a \textit{hackle}, an \textit{egg}, and a
\textit{whetstone}. A digital version of the collection is available
in the Dutch Folktale Database from the Meertens Institute (corpus
SINVSUNV.20E).\footnote{See \url{http://www.verhalenbank.nl}} Using a
single collection for our corpus presents us with a helpful
homogeneity with regard to the editor, length of the stories (which is
between X and Y words) and language use. On the other hand, the
collection displays diversity as well in the choice of the stories,
which entail fairy tales, legends, and sagas.

@@

This annotation resulted in 11542 animate tokens of 743 types, while
implicitly yielding 62926 inanimate tokens from 5011 unique inanimate
words. Because of our context-dependent approach, some words, such as
\textit{pancake} and \textit{egg}, occurred in both animate types as
inanimate types, because they were labeled as both animate and
inanimate in some cases in our corpus. It is telling that of the
animate tokens 4627 (40,1\%) were nouns, while only 6878 of the
inanimate tokens (11 \%) are nouns. This shows that being a noun is
already somewhat of an indication for animacy.
%hoeveel woorden hebben we eigenlijk die dubbel voorkomen, zoals pannekoek en ei? Leuk om te melden%
After tokenization with the tokenization module of the Python software
package Pattern \cite{smedt:12} we fed all stories to the
state of the art syntactic parser for Dutch, Alpino \cite{bouma:01}.
%dit kan iets uitgebreider eventueel%

@@

\subsection{Task description}

We formulate the problem of animacy detection as a classification
problem where the goal is to assign a label at word level, rather than
at lemma level. This label indicates whether the word is classified as
animate or inanimate.

@@

Animate words are far outnumbered by inanimate words in our collection
(see \S\ref{sec:data}). Reporting accuracy scores would therefore
provide skewed results, strongly favoring the majority category. The
relative rarity of animate words makes evaluation measures such as the
well-known $F1$-score more appropriate. For this reason, we report
on the precision, recall and $F1$-score~\cite{rijsbergen:79} for all experiments. As a
second measure of evaluation, we compute the Area Under the Precision
Recall Curve (AUC_{PR}) which also takes into account possible class
imbalance. Also, while in most of the literature on animacy detection
results are only presented for the classification of nouns or noun
phrases, we will, while reporting on nouns and noun phrases as well, additionally report on the
results for all words in a text.

@@

\subsubsection{Semantic Features} 
The most innovative feature we have
included in our model is concerned with semantic similarity. In his
\textit{Philophische Untersuchungen} Wittgenstein already suggests
that ``Die Bedeutung eines Wortes ist sein Gebrauch in der Sprache''
(PI 43). This is reflected by the well-known insight in computational linguistics that the
meaning of words can be approximated by comparing the linguistic
contexts in which words appear, is an implementation of this very
idea. The idea is very simple, yet powerful: words that often
co-appear with the same set of words, will have a more similar
meaning. Recently, there has been a lot of interest in procedures that
can automatically induce so-called `word embeddings' from large,
unannotated collections of texts
(e.g.~\cite{mikolov:13,pennington:14}). These models typically attempt
to learn a lower dimensional vector for each word in the vocabulary
which captures the typical co-occurrence patterns of the word in the
corpus. The similarity between words can then be approximated by
applying similarity metrics, such as the cosine metric, to these
vectors of word embeddings.

We have trained word embeddings with 300 dimensions using the popular
skip-gram architecture \cite{mikolov:13} on the Dutch corpus of COW
(COrpora from the Web).\footnote{The word embedding vectors have been
  published online at XXX.} COW is a collection of linguistically
processed web corpora for English, Dutch, Spanish, French, Swedish and
German \cite{schaefer:12}. The 2014 Dutch corpus contains 6.8 billion
word tokens. The idea behind using the word embeddings is that
similarities between animate words can be estimated by inspecting the
context in which they occur. From this follows, for example, that the word
embeddings of an animate word are more similar to those of other
animate words, as opposed to the embeddings of inanimate words.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{pca.pdf}
\caption{Two-dimensional PCA projection of the 300 dimensional word embedding
vectors for a number of animate and inanimate words. The horizontal line
illustrates the separabilty between the two classes in the first dimension.}
\label{fig:pca}
\end{figure}

To give an illustration of this idea, in Figure \ref{fig:pca} we depict a
two-dimensional Principle Component Analysis (PCA) projection of the 300 dimensional
word embedding vectors for a number of typically animate and typically inanimate
 words. The horizontal grey line in the plot illustrates the separability of
 the animate and inanimate words in the first dimension of the PCA projection.
 It is interesting to observe that the less animate entity \emph{ghost} is the
 one closest to all other inanimate entities. Likewise, words such as \emph{castle},
 \emph{house} or \emph{car} are often used in figurative language (metonomy),
 for example to refer to the people owning or living in the castle. Perhaps this
 ambiguous animacy position is reponsible for their position in the first dimension
 close to real animate entities.
%dit stukje hierboven kan nog iets worden uitgewerkt, eventueel met voorbeelden%
\subsection{Models}
We employ a Maximum Entropy classifier with L2 regularization as
implemented in \cite{sklearn}. In all experiments, we set the
regularization strength parameter $C$ to 1.

@@

Although our background corpus is sufficiently large to cover most
words in an unseen text, there will always be rare words for which we
do not have learned word embeddings. Therefore, in order to
effectively make use of the word embedding vectors, we need a way to
deal with out-of-vocabulary items. We adopt a simple strategy where we
make use of a primary classifier and a back-off classifier. For models
6 to 9, we augment each word with its corresponding 300 dimension word
embeddings vector. In the case of out-of-vocabulary words, we resort
to a back-off model that contains all features except the word
embeddings. For example, a model that makes use of words and word
embeddings, will make a prediction on the basis of the word features
alone. In case of the model that solely uses the embeddings (model 5),
the back-off classifier is represented by a majority vote classifier,
which classifies unseen words as inanimate.

@@

In Table \ref{tab:results-all} we present the results for all nine
models on the complete data set. For each model we report the
precision, recall, $F1$-score and Area Under the Precision-Recall
Curve (AUC_{PR}) for the animate words and the
inanimate words.
%AUC PR scores staan nu niet in de tabel, en worden schijnbaar niet gerapporteerd%
%positie van de tabellen is niet optimaal%
\begin{table}
\centering
\begin{tabular}{llrrr}
\toprule
features & class & Precision &  Recall & $F1$\\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.98 &    0.99 &    0.98 \\
                                                       & animate     &       0.93 &    0.89 &    0.91   \\
\multirow{2}{*}{word}                                  & inanimate     &       0.96 &    0.99 &    0.98 \\
                                                       & animate     &       0.94 &    0.78 &    0.85   \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.98 &    0.99 &    0.98 \\
                                                       & animate     &       0.94 &    0.90 &    0.91   \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.97 &    0.99 &    0.98 \\
                                                       & animate     &       0.94 &    0.86 &    0.89   \\
\rowcolor{Gray}                                        & inanimate     &       0.98 &    0.99 &    0.99 \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings}   & animate     &       0.94 &    0.91 &    0.93 \\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate    &       0.97 &    0.99 &    0.98  \\
                                                       & animate     &       0.94 &    0.86 &    0.90   \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.98 &    0.99 &    0.99 \\
                                                       & animate     &       0.94 &    0.91 &    0.93   \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.97 &    0.99 &    0.98 \\
                                                       & animate     &       0.94 &    0.86 &    0.90   \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.98 &    0.99 &    0.99 \\
                                                       & animate     &       0.94 &    0.92 &    0.93   \\
\bottomrule
\end{tabular}
\caption{Precision, Recall and $F1$-score for animate and inanimate classes per feature setting for all words.}
\label{tab:results-all}
\end{table}

@@

The model that only uses word embedding features is one of the best
performing models. This is a context insensitive model that operates
on the level of the vocabulary, which means that it will predict the
same outcome for each token of a particular word type. The high
precision and high recall show us that this model has acquired
knowledge about which words \emph{typically} group with animate
words and which with inanimate words. However, the models that combines
the word embeddings with the context sensitive features, such as word
$n$-grams or Part-of-Speech tags, outperform the context insensitive
model.

The best performance is achieved by the model that combines the word
features, part-of-speech tags and the word embeddings, which has an
$F1$-score of 0.93 on animate words and 0.99 on inanimate
words. Adding more features does not result in any more performance
gain.
%misschien hier dan nog een voorbeeld noemen, of verwijzen naar relevant cijfer in de tabel%
Table \ref{tab:results-noun} presents the results for all nouns and
names in the data set. The best performance is again achieved by the
model that combines the word features with the part-of-speech tags and
word embeddings, resulting in an $F1$-score of 0.92 for animate
instances and 0.95 for inanimate instances.
%moet in deze tabel ook AUC pr?%

\begin{table}
\centering
\begin{tabular}{llrrr}
\toprule
features & class &  Precision &  Recall & $F1$ \\
\midrule
\multirow{2}{*}{embeddings}                            & inanimate     &       0.90 &    0.96 &    0.92  \\
                                                       & animate       &       0.93 &    0.85 &    0.89  \\
\multirow{2}{*}{word}                                  & inanimate     &       0.78 &    0.98 &    0.87  \\
                                                       & animate       &       0.96 &    0.60 &    0.74  \\
\multirow{2}{*}{word + embeddings}                     & inanimate     &       0.90 &    0.97 &    0.93  \\
                                                       & animate       &       0.95 &    0.85 &    0.90  \\
\multirow{2}{*}{word + PoS}                            & inanimate     &       0.86 &    0.96 &    0.90  \\
                                                       & animate       &       0.93 &    0.78 &    0.84  \\
\rowcolor{Gray}                                        & inanimate     &       0.93 &    0.96 &    0.95  \\
\rowcolor{Gray}\multirow{-2}{*}{word + PoS + embeddings} & animate       &       0.95 &    0.90 &    0.92\\
\multirow{2}{*}{word + PoS + lemma}                    & inanimate     &       0.87 &    0.96 &    0.91  \\
                                                       & animate       &       0.94 &    0.80 &    0.86  \\
\multirow{2}{*}{word + PoS + lemma + embeddings}       & inanimate     &       0.93 &    0.96 &    0.94  \\
                                                       & animate       &       0.95 &    0.89 &    0.92  \\
\multirow{2}{*}{word + PoS + lemma + dep}              & inanimate     &       0.87 &    0.96 &    0.91  \\
                                                       & animate       &       0.93 &    0.80 &    0.86  \\
\multirow{2}{*}{word + PoS + lemma + dep + embeddings} & inanimate     &       0.93 &    0.96 &    0.95  \\
                                                       & animate       &       0.95 &    0.90 &    0.92  \\
\bottomrule
\end{tabular}
\caption{Precision, Recall, and $F1$ score for animate and inanimate classes
  per feature settings for all words tagged as noun.}
\label{tab:results-noun}
\end{table}

\section{Listing Characters in Stories}\label{sec:listing}
